---
title: "COVID19_DATA"
author: "Donal"
date: "2024-07-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

# Importing

## Set the import website and the filenames

```{r get_jhu_data}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", 
"time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)
```

## Read in the sets

```{r import_data, message=FALSE}
us_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
us_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
```

## Change the format

We want to change the date columns to rows to make the table more useful. Set Global cases = global cases, keep province,country, lat, long. also include two new columns one being date which takes the column names from the original table and the other column takes the values from the original table. We also remove lat and long at the end.

```{r tidy_global_data}
global_cases <- global_cases %>%
  pivot_longer(cols = - c("Province/State", "Country/Region", Lat, Long), 
              names_to = "date",
              values_to = "cases") %>%
              select(-c(Lat,Long))
              
global_deaths <- global_deaths %>%
  pivot_longer(cols = - c("Province/State", "Country/Region", Lat, Long), 
              names_to = "date",
              values_to = "deaths") %>%
              select(-c(Lat,Long))
```

## Merge the Global Cases and Global deaths and tidy the names

```{r merge_global_data}

global <- global_cases %>% 
    full_join(global_deaths) %>%
    rename(Country_Region = "Country/Region",
           Province_State = "Province/State") %>%
    mutate(date = mdy(date))
  
```

## Have a look at the summary stats and the dataset

```{r Summarize_N_Filter}
global <- global %>% filter(cases > 0)
summary(global)
global %>% filter(cases > 100000000)
```

## Do the same for the US datasets

```{r Same_for_US}
us_cases <- us_cases %>%
  pivot_longer(cols = - c(UID:Combined_Key), 
              names_to = "date",
              values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
              
us_deaths <- us_deaths %>%
  pivot_longer(cols = - c(UID:Population), 
              names_to = "date",
              values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))
  
us_tot <- us_cases %>% 
    full_join(us_deaths)
```

## Make a Combined Key for the Global Data

```{r CombKey_for_Glob}
global <- global %>%
  unite("Combined_key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```

## Take in the population data and merge it to global

```{r glob_pop_merge}
pop_tab <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")

pop_tab <- pop_tab %>%
  select(c(Country_Region, Province_State, Population))

global <- global %>%
  left_join(pop_tab)
```

## Group the US data by state and then group the entire US

```{r US_Grouping}
US_by_state <- us_tot %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_total <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
    select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
```

## Plot the Total US time series (ggplot - select the axes, geom_line - put a line in the chart, )

```{r US_Plotting}
US_total %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

state <- "Alabama"
US_by_state %>% filter(cases > 0, Province_State == state) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_sqrt() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ",state), y = NULL)
```

## Difference the time series data and plot it again

```{r US_Differences}
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_total <- US_total%>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_total %>% select(new_cases, new_deaths))

US_total %>%
  filter(new_cases > 0, new_deaths > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

US_by_state %>% filter(new_cases >= 0, new_deaths >= 0, Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_sqrt() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ",state), y = NULL)
```

## Cross State comparisons

```{r US_State_totals}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population  = max(Population),
            cases_per_thou = 1000* cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

US_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

```

## Modelling your data - look at the max and min and do a linear regression

```{r Simple_model}

mod <- lm(deaths_per_thou ~ cases_per_thou, data=US_state_totals)
summary(mod)

US_state_totals %>% slice_min(cases_per_thou)
US_state_totals %>% slice_max(cases_per_thou)

US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred %>% ggplot() +
  geom_point(aes(x= cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x= cases_per_thou, y = pred), color = "red")

US_tot_w_pred <- US_tot_w_pred %>% mutate(deaths_per_case = deaths/cases)

US_tot_w_pred %>% filter(population> mean(population))%>%
  ggplot() +
  scale_y_log10() +
  geom_point(aes(x = Province_State, y = deaths, color="deaths"))+
  geom_point(aes(x = Province_State, y = cases, color="cases")) +
  theme(axis.text.x = element_text(angle = 90))


quanties <- unname(quantile(US_tot_w_pred$population))

US_tot_w_cat <- US_tot_w_pred %>% 
  mutate(catty = ifelse(population <= quanties[2], "small",ifelse(population > quanties[4],"large","med"))) 

US_tot_w_cat %>% ggplot(aes(catty,deaths_per_thou)) +
  geom_boxplot() +
  geom_point(aes(size = cases, colour = catty, alpha = .5)) +
  labs(title = "Deaths per 1,000 by state size") 
  

```

## Sources of bias

Some of the sources of bias that may be in this report include: 

* Half way through 2023 there appear to be a lot of days where no covid cases or deaths are recorded in Alabama followed by a day when thousands were. Did that state stop reporting cases and deaths on a daily basis? Maybe a moving average over 5 days would be better for analysis.

* The incentive for people to report that they were covid positive probably lessened as 2022/23 continued. 

* I assumed that the link between new case numbers and deaths would be broken by the vaccine and that more populated states would have a higher death and caseload per thousand.

Again it is always important to understand the context in which your data was collected.

## Conclusion

COVID deaths plateaued around the end of 2022 and into 2023 and for both new cases and new deaths; the inter day variation between quantities increased which might indicate that something is happening with the data in that period.

When I modeled the deaths on the number of cases on a state level, the linear regression was not a good model based on the p-value and the plot. Underlying demographics, state policy and vaccine uptake probably broke the relationship but that would require further investigation.

Population appeared to be linked to the deaths per thousand based on the boxplot and I couldn't spot any outliers when it came to deaths and cases in the cases/deaths plot of the most populous states.
